---
title: "Dry biomass of freshwater invertebrates using InvTraitR"
output: rmarkdown::html_vignette
author: "James Hagan"
vignette: >
  %\VignetteIndexEntry{Dry biomass of freshwater invertebrates using InvTraitR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(InvTraitR)
```

```{r}
# load relevant libraries
library(dplyr)
library(readr)
library(tidyr)
library(ggplot2)
library(assertthat)
library(testthat)
```
To see how we can use *InvTraitR* to obtain estimates of the dry biomass of freshwater invertebrates using taxon names and geographical coordinates, we will start by loading some relevant packages and some test data. The test data consist of invertebrate taxa collected in six temporary ponds in a nature reserve in Belgium. In addition, the data consist of latitude-longitude coordinates for the samples as well as measured body lengths for all the taxa.

```{r}
# load the test data
pond_df <- readr::read_csv("https://raw.githubusercontent.com/haganjam/InvTraitR/main/database/InvTraitR_vignette_data.csv")
head(pond_df)

# how many unique taxa are there?
length(unique(pond_df$target_taxon))

# check the summary statistics
summary(pond_df)
```

We want to use the body length data to estimate the dry biomass for each of these samples using length-dry biomass allometric equations. The *InvTraitR* package houses a database of more than 350 length-dry biomass allometric equations that represent most major groups of freshwater invertebrates globally. We can use *InvTraitR* to search this database for appropriate length-dry biomass equations for the samples in our test data (`pond_df`).

To do this, we have two options both of which rely on the `get_trait_from_taxon()` function. This function (see below) takes a data.frame with at least five columns. These columns are the taxon name for which dry biomass data are required, the life stage of the taxon, the latitude and longitude coordinates from which the taxon was sampled and body size of the taxon:

```{r eval=FALSE}
get_trait_from_taxon(
    data,                   # data.frame with at least five columns: target taxon, life stage, latitude (dd), longitude (dd) and body size (mm) if trait == "equation"
    target_taxon,           # character string with the column name containing the taxon names
    life_stage,             # character string with the column name containing the life stages
    latitude_dd,            # character string with the column name containing the latitude in decimal degrees
    longitude_dd,           # character string with the column name containing the longitude in decimal degrees
    body_size,              # character string with the column name containing the body size data if trait = "equation"
    workflow = "workflow2", # options are "workflow1" or "workflow2" (default = "workflow2)
    max_tax_dist = 3,       # maximum taxonomic distance acceptable between the target and the taxa in the database (default = 3)
    trait = "equation",     # trait to be searched for (default = "equation")
    gen_sp_dist = 0.5       # taxonomic distance between a genus and a species(default = 0.5)
)
```

### workflow1

The first option is *workflow1*. If *workflow1* is chosen, `get_trait_from_taxon()` will search the database and, for each sample in your dataset, it will output all equations that are within the maximum taxonomic distance (*max_tax_dist*) set by the user. We can then use our own discretion to choose which of these equations we would like to use.

Let's see how this works:

```{r}
# we specify the dol_df into the function
pond_equ1 <- get_trait_from_taxon(
  data = pond_df,
  target_taxon = "target_taxon",
  life_stage = "life_stage",
  latitude_dd = "lat",
  longitude_dd = "lon",
  body_size = "length_mm", 
  workflow = "workflow1",
  trait = "equation",
  max_tax_dist = 3.5,
  gen_sp_dist = 0.25
)
```


The output is a list of length two with two different data.frames: the data with relevant equations (*data*) and a data.frame with information on how we chose the equations (*decision_df*).

```{r}
# look at the first few rows of both the output data.frames in the list
head(pond_equ1$data)
head(pond_equ1$decision_data)
```

We'll start by looking at the outputted data with the equations. To do this, let's simply look at the first row of the raw data. 

```{r}
# check the first row of the data that we fed into the function
print(pond_df[1,])
```

Then, let us compare this with what the function outputs:

```{r}
# check the first row of the output data with the equations
print(dplyr::filter(pond_equ1$data, row == 1))
```

In this output, we have several rows for each input row (these are labelled with the row column i.e. input row 1 is labelled as '1' in the row column). For each row from the input data, there are several rows in this output data. These different rows correspond to different equations from the database that are within the maximum taxonomic distance that we set (i.e. 3.5).

The idea with *workflow1* is for us to now use our discretion to choose appropriate equations. To look in more detail at the output, we can also use the *View()* function:

```{r eval=FALSE}
View(dplyr::filter(pond_equ1$data, row == 1))
```

The most important columns to consider are the *tax_distance* column, the *body_size_range_match* column and the *life_stage_match* column to make an initial decision. We want low taxonomic distances, we want the length of our target data to be within the body size range of the equation in the database (which is specified by the *body_size_range_match* column) and, very importantly, we want to the life-stage of our target to match the life-stage of the equation. There is only one equation that matches the life-stage (id = 169). Given that the target taxon (*Acilius sulcatus*) is a beetle, life-stage is crucially important. The problem, however, with id = 169 is that the body size range does not match. Indeed, the equation was developed for body sizes between 1.4 and 5.2 mm (*db_min_body_size_mm* and *db_min_body_size_mm* columns) and the length in our data is 16.3 mm. Thus, using this equation would be a considerable extrapolation. Given the non-linearity of many allometric equtaions, this may be a problem.

Nonetheless, because the other equations within the appropriate taxonomic do not have the appropriate life-stage, this is the best that we can do and use our expert knowledge to check that the biomass value that we obtain is reasonable for this taxon (e.g. using other literature, for example). These choices are, of course, subjective to a certain degree. Choosing between id = 167 or id = 168 is at some level subjective. The point of *workflow1* is to provide the relevant information to make informed decisions.

Based on this information, we will choose id = 169 because it is the only equation with the appropriate life-stage information.

```{r}
ex1 <-
  pond_equ1$data |>
  dplyr::filter(row == 1, id == 169) |>
  dplyr::select(row, target_taxon, life_stage, lat, lon, length_mm,
                equation_form, log_base, a, b, lm_correction, dry_biomass_scale)
```

Using these data, we can simply calculate the expected dry biomass based on the chosen equation as follows:

```{r}
# use with so we can directly reference the variables
ex1_biomass_mg <- with(ex1, {
  # calculate the raw prediction on the log-scale
  x <- a + (b * logb(x = length_mm, base = log_base))

  # convert to the natural scale
  x <- (log_base^x)

  # apply the correction factor and scaling factor
  dry_biomass_mg <- dry_biomass_scale * ifelse(
    !is.na(lm_correction),
    x * lm_correction,
    x
  )

  return(dry_biomass_mg)
})

print(paste0("Expected dry biomass (mg) = ", round(ex1_biomass_mg, 2)))
```

Checking the literature, I found a that ca. 50 mg is not atypical for a Dysticid diving beetle adult which provides confidence in this estimate (Klecka and Boukal 2013, Journal of Animal Ecology).

We've developed a helper function to automatically perform these calculations based on the following variables in a data.frame.

### workflow2

The second option for obtaining dry biomass estimates for a set of taxa is workflow2 which is our automated workflow for selecting equations. Let's see how this works:

```{r}
# we specify the dol_df into the function and select workflow2
pond_equ2 <- get_trait_from_taxon(
  data = pond_df,
  target_taxon = "target_taxon",
  life_stage = "life_stage",
  latitude_dd = "lat",
  longitude_dd = "lon",
  body_size = "length_mm", 
  workflow = "workflow2",
  trait = "equation",
  max_tax_dist = 3.5,
  gen_sp_dist = 0.25
)
```

Unlike with *workflow1*, when we open the output data, it has the same number of rows as the input data (i.e. *pond_df*). That is because *InvTraitR* automatically selected an appropriate equation for each row where an appropriate equation was available in the database. Moreover, *InvTraitR* directly calculated the expected dry biomass (see *dry_biomass_mg* column) For example, let's look at the first row:

```{r}
# check the first row of the output data with the equations
print(dplyr::filter(pond_equ2$data, row == 1))
```

In this case, all the columns are filled with NA values. This is because there is no appropriate equation for this taxon according to the InvTraitR *workflow2* criteria. For an equation to be selected with *workflow2*, it has to meet three conditions:

1. The taxonomic distance must be less than the maximum taxonomic taxonomic distance (in this case 3.5 as we set it)

2. The target body size must be within the body size range of the equation

3. The life-stage of the target taxon and the equation must match

As was clear from looking at this example with *workflow1*, there were no equations that met all three of these criteria. Therefore, we ended up using an equation where the target body size was not within the body size range of the equation,

Indeed, if we look at the whole output, many of the rows do not have equations which means that *InvTraitR* was unable to find appropriate equations in the database based on these three criteria. 

```{r eval=FALSE}
View(pond_equ2$data)
```

In fact, in this dataset, *InvTraitR* was only able to find equations for 25 % of the data points and 29 % of the unique taxa.

```{r}
# proportion of data points for which appropriate equations were found
sum(!is.na(pond_equ2$data$id))/nrow(pond_equ2$data)

# proportion of unique taxa for which appropriate equations were found
n <-
  dplyr::filter(pond_equ2$data, !is.na(id)) |>
  dplyr::pull(target_taxon) |>
  unique() |>
  length()

n / length(unique(pond_equ2$data$target_taxon))
```
If we want to get more insight into why *InvTraitR* was unable to find appropriate equations for the different rows, we can look at the second output which is the *decision_data*. For each row, it shows all the equations that were considered and then in the column called *explanation*, it explains why a given equation was not chosen.

```{r eval=FALSE}
View(pond_equ2$decision_data)
```

However, in this case, we specified certain levels of maximum taxonomic distance in the function. What happens if we relax this and increase the acceptable maximum taxonomic distance and increase it to 4 (instead of 3.5)?

```{r}
# we specify the dol_df into the function and select workflow2
pond_equ2 <- get_trait_from_taxon(
  data = pond_df,
  target_taxon = "target_taxon",
  life_stage = "life_stage",
  latitude_dd = "lat",
  longitude_dd = "lon",
  body_size = "length_mm", 
  workflow = "workflow2",
  trait = "equation",
  max_tax_dist = 4,
  gen_sp_dist = 0.25
)
```

Do we get a higher proportion of datapoint and taxa with appropriate equations?

```{r}
# proportion of data points for which appropriate equations were found
sum(!is.na(pond_equ2$data$id))/nrow(pond_equ2$data)

# proportion of unique taxa for which appropriate equations were found
n <-
  dplyr::filter(pond_equ2$data, !is.na(id)) |>
  dplyr::pull(target_taxon) |>
  unique() |>
  length()

n / length(unique(pond_equ2$data$target_taxon))
```
In this case, we do not. If this occurs, a user can see which datapoints appropriate equations were not given for and then use the decision data to see if there are appropriate equations for taxa that *InvTraitR* missed. Otherwise, if no good equation can be found, one might have to use order-level equations instead which are available in many research papers (e.g. Benke et al. 1999).
